---
title: "Final Exercise"
author: "Perry S"
format:
  html:
    code-link: true
code-copy: true
execute: 
  message: false
  warning: false
---

```{r}
#| include: false

library(tidyverse)
library(here)

df_wq <- read_csv(here("data/WQ_P8D7.csv"))
```

## Final Exercise

Using pipes, modify the `df_wq` data frame as so:

-   assign it to a new object named `df_new`

-   filter `Date` to 2021 data only

-   select only the `Station`, `Date`, `Chla`, and `Pheophytin` columns

-   mutate `Station` in-place so it's a factor column

**Challenge**:

Modify the `df_new` data frame as so:

-   mutate `Chla` so it's +20 specifically for the *case when* the month is April *and* the station is P8

    -   *Hint: April is the 4th month*

-   assign it to the object `df_new` (ie. the same name)

    -   can you access the "original" `df_new` now? what happens if you re-run the code?

```{r}
#| code-fold: true
#| eval: false

df_new <- df_wq %>%
  filter(Date > '2020-12-31' & Date < '2022-12-31') %>% # can also use lubridate
  select(c(Station, Date, Chla, Pheophytin)) %>%
  mutate(Station = factor(Station))

glimpse(df_new)
```

```{r}
#| code-fold: true
#| eval: false

df_new <- df_new %>%
  mutate(
    Chla = case_when(
      month(Date) == 04 & Station == 'P8' ~ Chla + 20,
      TRUE ~ Chla
    )
  )

df_new
```
