---
title: "Mutate Data"
author: "Perry S"
format:
  html:
    code-link: true
code-copy: true
execute: 
  message: false
  warning: false
---

```{r}
#| include: false

library(tidyverse)
library(here)

df_wq <- read_csv("data/WQ_P8D7.csv")
```

## Introduction

Sometimes, we want to modify existing columns values or even add new ones. This is where the `mutate` function from the `dplyr` package comes in.

## Add New Columns

With `mutate`, it's simple to add new columns. First, let's look at our current ones:

```{r}
glimpse(df_wq)
```

\
Now let's add a new column, *Lab*, where every value is "BSA":

```{r}
df_wq %>%
  mutate(Lab = 'BSA') %>%
  glimpse()
```

We can even add columns that are combinations of other ones:

*add two numeric columns together*

```{r}
df_wq %>%
  mutate(ChlPheo = Chla + Pheophytin) %>%
  select(Station, Date, Chla, Pheophytin, ChlPheo) %>%
  head()
```

*concatenate two character columns*

```{r}
df_wq %>%
  mutate(StatDate = paste(Station, Date, sep = '___')) %>%
  select(Station, Date, StatDate) %>%
  head()
```

::: callout-note
## Relocate Columns

Sometimes, we want to re-arrange our column order. For example, with our new `Lab` column, we might want it in the beginning with the other metadata. Here, we use the `relocate` function:

```{r}
df_wq %>%
  mutate(Lab = 'BSA') %>%
  relocate(Lab, .after = Date) %>%
  head()
```
:::

## Modify Existing Columns

We can also use mutate to modify existing columns. For example, say we want to add 20 to every *Chla* value:

```{r}
df_wq %>%
  mutate(Chla = Chla+20)
```

We can even change data types! This is useful if, say, you have character column that should be numeric, or a character column that should be factor. We do this by using the `as.numeric` and `as.factor` functions (`as.character` is also an option):

```{r}
df_wq %>%
  mutate(Chla = as.numeric(Chla),
         Station = as.factor(Station)) %>%
  select(Station, Chla) %>%
  str()
```

## `case_when`

Sometimes you only want to change certain values in a column. For example, maybe you want to add 20 to `chla` values, but only for rows where the station is P8. You can do this using the `case_when()` function for this.

`case_when()` lets you assign new values based on conditions. It takes each row and sequentially compared it to each condition, using the value for the first one that is true. The format looks like this:

```         
case_when(
  condition1 ~ value1,
  condition2 ~ value2,
  TRUE ~ default_value
  )
```

Since every row must get a value, it’s important to end with `TRUE ~ result` to handle all remaining cases.

```{r}
df_wq %>%
  mutate(
    Chla =
      case_when(Station == 'P8' ~ Chla + 20, # if station is P8, add 20
              TRUE ~ Chla) # else, keep as previous value
  ) %>%
  select(c(Station:Date), Chla)
```

## Rename

Sometimes, we want to rename our columns (eg. if working with multiple data frames, we want the columns to be standardized). We can do this using the `rename` function from the `dplyr` package:

```{r}
df_wq %>%
  rename(StationCode = Station)
```
