---
title: "Mutate Data"
author: "S Perry"
format:
  html:
    code-link: true
code-copy: true
execute: 
  message: false
  warning: false
---

```{r}
#| include: false

library(tidyverse)
library(here)

df_wq <- read_csv("data/WQ_P8D7.csv")
```

## Introduction

Sometimes, we want to modify existing columns values or even add new ones. This is where the `mutate` function from the `dplyr` package comes in.

## Add New Columns

With `mutate`, it's simple to add new columns. First, let's look at our current ones:

```{r}
glimpse(df_wq)
```

\
Now let's add a new column, *Lab*, where every value is "BSA":

```{r}
df_wq %>%
  mutate(Lab = 'BSA') %>%
  glimpse()
```

We can even add columns that are combinations of other ones:

*add two numeric columns together*

```{r}
df_wq %>%
  mutate(ChlPheo = Chla + Pheophytin) %>%
  select(Station, Date, Chla, Pheophytin, ChlPheo) %>%
  head()
```

*concatenate two character columns*

```{r}
df_wq %>%
  mutate(StatDate = paste(Station, Date, sep = '___')) %>%
  select(Station, Date, StatDate) %>%
  head()
```

::: callout-note
## Relocate Columns

Sometimes, we want to re-arrange our column order. For example, with our new `Lab` column, we might want it in the beginning with the other metadata. Here, we use the `relocate` function:

```{r}
df_wq %>%
  mutate(Lab = 'BSA') %>%
  relocate(Lab, .after = Date) %>%
  head()
```
:::

## Modify Existing Columns

We can also use mutate to modify existing columns. For example, say we want to add 20 to every *Chla* value:

```{r}
df_wq %>%
  mutate(Chla = Chla+20)
```

We can even change data types! This is useful if, say, you have character column that should be numeric, or a character column that should be factor. We do this by using the `as.numeric` and `as.factor` functions (`as.character` is also an option):

```{r}
df_wq %>%
  mutate(Chla = as.numeric(Chla),
         Station = as.factor(Station)) %>%
  select(Station, Chla) %>%
  str()
```

## Formatting Dates

A particularly useful application for mutate is to properly format dates. Remember we used `lubridate` to extract information, like year, from our dates:

```{r}
year(df_wq$Date) %>% head()
```

Will these functions work if Date is in a different format?

Clearly, this is incorrect; to R, the only correct date format is Year-Month-Day (with a 4 digit year). We can convert dates using the `as.Date` function in R. This can also be used in the `mutate` function.

To do this, we need to specify the format the other date is in. A quick cheat sheet:

- `%Y`: year (4 digit)
- `%y`: year (2 digit)
- `%m`: month (numeric)
- `%B%`: month (full name)
- `%b%`: month (abbreviation)
- `%d`: day
- `%H`: hour
- `%M`: minute
- `%S`: second

```{r}
test_date <- '04/05/2021'

new_date <- as.Date(test_date, format = '%m/%d/%Y')

new_date
class(new_date)
```

### Exercise

Convert `April 5 2021` to a date character

```{r}
test_date2 <- 'April 5 2021'

as.Date(test_date2, format = '%B %d %Y')
```

## `case_when`

Sometimes you only want to change certain values in a column. For example, maybe you want to add 20 to `chla` values, but only for rows where the station is P8. You can do this using the `case_when()` function for this.

`case_when()` lets you assign new values based on conditions. It takes each row and sequentially compared it to each condition, using the value for the first one that is true. The format looks like this:

```         
case_when(
  condition1 ~ value1,
  condition2 ~ value2,
  TRUE ~ default_value
  )
```

Since every row must get a value, it’s important to end with `TRUE ~ result` to handle all remaining cases.

To go back to our example, the code would look like this:

```{r}
df_wq %>%
  mutate(
    Chla =
      case_when(Station == 'P8' ~ Chla + 20, # if station is P8, add 20
              TRUE ~ Chla) # else, keep as previous value
  ) %>%
  select(c(Station:Date), Chla)
```

Note that `case_when` is nested in the `mutate` function and that we had to specify which column we wanted to manipulate (in this case, Chla). As you may assume, since it's part of mutate, this means we can also create new columns via this process.

#### Exercise

Now its your turn to try out these summarizing functions we just learned about.

1)  Use `case_when` to multiple Pheophytin values by 2 for station D7.

2)  Use `case_when` (nested in `mutate`) to create a new column, `DON_ND`, where the value is 0 when when DON is `NA` (ie. a non-detect)\
    **HINT:** we can use `is.na` to check if a column is NA

Click below for the answer when you are done!

```{r}
#| code-fold: true
#| eval: false
#| 
df_wq %>%
  mutate(
    Test = 
      case_when(is.na(DON) ~ 0,
                TRUE ~ DON)
  )
```

## Rename

Sometimes, we want to rename our columns (eg. if working with multiple data frames, we want the columns to be standardized). We can do this using the `rename` function from the `dplyr` package:

```{r}
df_wq %>%
  rename(StationCode = Station)
```
