---
title: "Import Data"
author: "Perry S"
format:
  html:
    code-link: true
code-copy: true
execute: 
  message: false
  warning: false
---

## Load Packages

One of the most important *packages* for this task is `tidyverse`, which we learned about yesterday. We'll import it here:

```{r}
library(tidyverse)
```

## Read in Data

Next, we must import our data. The common function for this is `read_csv` from `readr` (which is nested in `tidyverse`). For our demonstrations, we will use the **WQ_P8D7.csv** file housed in the **data** folder.

When importing data, we must specify the *filepath* where it's housed. There are multiple ways to do this. We could hard-code in the path, which is called the **absolute path**:

```{r}
#| error: true

df_wq <- read_csv('C:/R/IntrotoR/data/WQ_P8D7.csv')
```

However, this code is very specific and will break when used on other computers. If we instead house data in a **Project**, we can make use of **relative filepaths**. These are ideal because anyone who uses the Project can run the code:

```{r}
df_wq <- read_csv('data/WQ_P8D7.csv')
```

**If you received an error here**, this is because your Rmd files are not being evaluated at the Project level. One way to fix this is to change your options in `Tools` \> `Global Options` \> `R Markdown` \> `evaluate chunks in directory` to `Project`.

You can also use the `here` package instead, which automatically evaluates at the Project level:

```{r}
library(here)

df_wq <- read_csv(here('data/WQ_P8D7.csv'))
```

------------------------------------------------------------------------

We now have a data frame *object* called `df_wq`. We can use `head` to see what this data frame looks like:

```{r}
head(df_wq)
```

And `glimpse` to see information about the columns:

```{r}
glimpse(df_wq)
```
