---
title: "Day 4 - Summarizing and Reshaping Data, Course Recap"
author: "Dave Bosworth"
format: html
---

```{r}
library(tidyverse)
library(here)
```

```{r}
df_wq <- read_csv(here("data/WQ_P8D7.csv"))
```

# Summarizing Data

## Basic Summarizing

```{r}
# Simple overall summary statistics of entire data frame
summary(df_wq)

# Summary of one column
summary(df_wq$Chla)
```

```{r}
# Calculate overall mean of one column
df_wq %>% summarize(mean_Chla = mean(Chla))

# Calculate overall mean of multiple columns
df_wq %>% summarize(mean_Chla = mean(Chla), mean_Pheo = mean(Pheophytin))
```

## Grouping

```{r}
# Group by Station
df_wq %>% group_by(Station) %>% summarize(mean_Chla = mean(Chla))
```

```{r}
# Calculate more than one summary statistic within `summarize`
df_wq %>% 
  group_by(Station) %>% 
  summarize(
    min_Chla = min(Chla),
    mean_Chla = mean(Chla),
    median_Chla = median(Chla),
    max_Chla = max(Chla),
    sd_Chla = sd(Chla)
  )
```

It would be good to show how to group by more than one variable. This may be a good example of using pivot_longer to create a parameter variable to group by. We could also add an additional categorical variable to the existing data set - maybe season, year (add a year of data), etc.

# Reshaping Data

## Basics of Data Structure

-   "long" vs. "wide" structure
-   show examples

## Converting between Data Structures

### "Wide" to "Long"

### "Long" to "Wide"

# Joins and Binds

## Differences between them

## Bind rows

## Join basics

## Left join

# Final Course Exercise
